Extended Abstract

### 2021-05-11 paper/paper.tex
Another go at the introduction

\begin{abstract}
  The linear \picalc{} is a well-known substructural type system for the \picalc{}:
  it is a common target for the encoding of session-typed languages \cite{DardhaGS17},
  and it in direct correspondence with classical linear logic \todo{cite}.

  Type inference for the linear \picalc{} comes with its own challenges:
  for one, processes themselves are untyped, and thus typing constraints only apply to the typing contexts;
  moreover, typing constraints on usage annotations \emph{cannot always be eagerly solved in open processes}.

  We approach the first challenge using \emph{co-contextual type inference} \cite{ErdwegBKKM15}, where processes are traversed bottom-up and constraints are collected (and when possible eagerly solved) and merged on context splits.
  We approach the latter challenge by allowing the resolution of some of the constraints to be deferred.
  We state soundness and completeness theorems that take this deferral of constraint resolution into account.
  We state a progress (?) theorem that shows that only those constraints that need to be deferred are deferred.
  \todo{be more precise}
  We are working on mechanising our work in Agda.
\end{abstract}

@@

We define an untyped but well scoped \picalc{} using type-level de Bruijn \cite{deBruijn72} indices \autoref{syntax}.
We define a small expression language within the process syntax to handle composite sum and product types.
On top of the well-scoped syntax we provide the standard linear type-system using context-splits \autoref{type-system}.
The type system is indexed by a typing context that admits unsolved metavariables.
These unsolved metavariables are however well kinded: either types or usage annotations.

\todo{insert example showing why we need deferral}

We split typing inference into constraint collection and constraint solving \autoref{inference}.
We start by defining constraints on types and usage annotations \autoref{constraints}
We collect constraints bottom-up, creating extra \emph{merging} constraints on context splits.
This allows type-inference to be easily parallelisable.
We postulate soundness and completeness theorems that allow for constraint solving to be deferred \autoref{inference-theorems}.

>>>>

\begin{abstract}
  The linear \picalc{} is a well-known substructural type system for the \picalc{}:
  it is a common target for the encoding of session-typed languages \cite{DardhaGS17}
  and in direct correspondence with classical linear logic \todo{cite}.

  For the type checking of the linear \picalc{} to be decidable the newly created channels must have their types synthesized.
  Inferring these types can prove challenging, as not all their typing constraints are local.
  Moreover, some typing constraints cannot be eagerly solved and need to be deferred.

  Following \cite{Padovani15}, we split this problem in two: first we provide a \emph{co-contextual type inference} \cite{ErdwegBKKM15} algorithm that traverses processes bottom-up collecting constraints and merging them on context splits; then we solve some of these constraints using well-known unification algorithms \cite{McBride03}, while we defer the resolution of those constraints that lack enough information.
  We provide clear soundness and completeness theorems for both these stages, and show that only those constraints that lack enough information are deferred.
  The totality of this work is being mechanised in Agda.
\end{abstract}

@@

Consider a linear \picalc{} process with a free channel variable $x$ and wherein we create a new channel $y$, use $y$ to transmit, then send $y$ away over $x$ and terminate.
Unless the payload type of $x$ is determined elsewhere, in this process the usage annotations for the channel $y$ are polymorphic, yet the output capability on $y$ requires a lower bound of multiplicity $\tone$ (since the process uses it to transmit).
While some of the constraints on $x$ and $y$ can be eagerly solved, others are non-local and need to be deferred.

We approach this problem by breaking down typing inference into the collection of constraints \autoref{inference} and their (partial) resolution \autoref{constraint-resolution}.
We start defining an untyped but well scoped \picalc{} using type-level de Bruijn \cite{deBruijn72} indices \autoref{syntax}.
The syntax for processes contains a small expression syntax that handles composite sum and product types.
On top of the well-scoped syntax we define a standard linear type system using context-splits \autoref{type-system}.

Constraints are defined on types and usage annotations alike \autoref{constraints}.
We use a common metavariable context for both types and usage annotations.
The metavariable context is kinded, ans substitution is defined so that it respects kindedness.
In \autoref{inference} constraints are collected bottom-up, making type-inference easily parallelisable.
Context splits are solved by creating a fresh extra context and constraints that \emph{merge} the three.
We prove a soundness theorem and postulate a completeness theorem, both allowing for constraint solving to be deferred \autoref{inference-theorems}.

Constraints of the form $\eqconstr{x}{y}$ are solved by unification using a kinded version of \cite{McBride03}.
Constraints of the form $\eqconstr{x}{y}{z}$ are solved recursively until a base case is reached, when they are either admitted as valid or deferred.
We postulate soundness and completeness theorems and show that only those constraints without a unique solution are deferred.
